---
title: "lab2"
author: "Shaibu Abdullateef Topa"
date: "2025-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load Packages
```{r}
library(caret)
library(datasets)
library(partykit)
library(e1071)
library(C50)
library(RWeka)
library(rattle)
```
#Show datasets
```{r}
library(help="datasets")
```

#Show partykit
```{r}
library(help="datasets")
```


#Show data WeatherPlay
```{r}
weatherData <- WeatherPlay
view(weatherData)
```

#Dimension of Weather data
```{r}
dim(weatherData)
```
#Attributes of Weather data
```{r}
attributes(weatherData)
```
#Number of instances of Weather data
```{r}
nrow(weatherData)
```
#Tell R you will be using the Weather dataset by typing
```{r}
WeatherPlay
```


#Algorithm J48 to Train
```{r}
set.seed(1)
j48Tree <- train(
  play ~ .,
  data = WeatherPlay,
  method = "J48",
  trControl = trainControl(method = "cv", number = 5)
)
```


#Show result
```{r}
summary(j48Tree$finalModel)
```
#Plot result
```{r}
plot(j48Tree$finalModel)
```

#Exercises 1 - 1 Using C5 Algorithm
```{r}
set.seed(1)
c5Tree <- train(
  play ~.,
  data = WeatherPlay,
  method = "C5.0Tree",
  trControl = trainControl( method = "cv", number = 5)
)

summary(c5Tree)

c5Treee <- C5.0(play ~., data = WeatherPlay)
plot(c5Treee)
```



#Using C5.0Rules
```{r}
set.seed(1)
c5Rules <- train(
  play ~.,
  data = WeatherPlay,
  method = "C5.0Rules",
  trControl = trainControl( method = "cv", number = 5)
)

summary(c5Rules$finalModel)
```


```{r}
data(segmentationData)
dim(segmentationData)
```


```{r}
view(segmentationData)
```

#Remove attribute cell as it is not useful
```{r}
segmentationData$Cell <- NULL
```

# Split into training and testing datasets
```{r}
STrain <- subset(segmentationData, Case == "Train")
STest  <- subset(segmentationData, Case == "Test")
```

#remove Case attribute from both dataset 
```{r}
STrain$Case <- NULL
STest$Case  <- NULL
```

#Build a model using the train set
```{r}
set.seed(1)
STree <- train(
  Class ~ .,
  data = STrain,
  method = "J48",
  trControl = trainControl(method = "cv", number = 5)
)

summary(STree$finalModel)
plot(STree$finalModel)
```





#To test the model use the predict function. 
```{r}
TestRes <- predict(STree, newdata = STest, type="raw")
confusionMatrix(TestRes, STest$Class)
```
#Apply J48, C5.0Trees and C5.0Rules to ContactLenses 
```{r}
contactLens <- read.csv("ContactLens.csv")

set.seed(1)

LensJ48 <- train(
  ContactLens ~ .,
  data = contactLens,
  method = "J48",
  trControl = trainControl(method = "cv", 5)
)

summary(LensJ48$finalModel)
plot(LensJ48$finalModel)
```
#Apply J48, C5.0Trees and C5.0Rules to ContactLenses 
```{r}
contactLens <- read.csv("ContactLens.csv")

set.seed(1)

LensC5Trees <- train(
  ContactLens ~ .,
  data = contactLens,
  method = "C5.0Tree",
  trControl = trainControl(method = "cv", 5)
)

summary(LensC5Trees$finalModel)

```
#Apply J48, C5.0Trees and C5.0Rules to ContactLenses 
```{r}
contactLens <- read.csv("ContactLens.csv")

set.seed(1)

LensC5Rules <- train(
  ContactLens ~ .,
  data = contactLens,
  method = "C5.0Rules",
  trControl = trainControl(method = "cv", 5)
)

summary(LensC5Rules$finalModel)
```
```{r}
diabetes <- read.csv("diabetes.csv")

set.seed(1)

rPart <- train(
  Outcome ~ .,
  data = diabetes,
  method = "rpart",
  trControl = trainControl(method = "cv", 5)
)

summary(rPart$finalModel)
fancyRpartPlot(rPart$finalModel)
```
#Download file heart-c.arff from CampusMoodle.
#The class is num.
#Attempt to apply J48 to this dataset.

```{r}
heart <- read.arff("heart.c.arff")
set.seed(1)
# heartJ48 <- train(
#   num ~ .,
#   data = heart,
#   method = "J48",
#   trControl = trainControl(method = "cv", 5)
# )

# summary(heartJ48$finalModel)
# The above fails (ERROR: na.fail.default) bcoz the dataset contains missing values

noNAHeart <- na.omit(heart)

noNAHeart$num <- droplevels(noNAHeart$num)

heartJ48 <- train(
  num ~ .,
  data = noNAHeart,
  method = "J48",
  trControl = trainControl(method = "cv", number = 5)
)

summary(heartJ48$finalModel)
plot(heartJ48$finalModel)
```
