---
title: "Lab Exercise One"
author: "Shaibu Abdullateef Topa"
date: "2025-02-03"
output: html_document
---

##Lab Exercise 1
power of all numbers between -8 and 13

```{r}
numbers <- c(-8:13)
power_of_numbers <- numbers ^ 2
power_of_numbers
```
##Lab Exercise 2
Data Frame for courses
```{r}
course <-  c("Data Science", "Cyber Security", "IT", "Digital Media")
module1 <- c("analysis", "ethical hacking", "databases", "3D animation")
credits1 <- c(15,15,20,30)
module2 <- c("project", "forensics", "java programming", "perception")
credits2 <- c(45,15,20,12)
courses <- data.frame(course, module1, credits1, module2, credits2)
head(courses)
```
##Lab Exercise 3
Extend courses with a row with the following information
```{r}
newrow <- data.frame("Networks", "CISCO CCNA", 25, "Operating systems", 40)
names(newrow) <- names(courses)
courses <- rbind(courses, newrow)
head(courses)
```
##Lab Exercise 4
Extend courses with a column total which contains the total number of credits for a course.
```{r}
courses$total <- courses$credits1 + courses$credits2
head(courses)
```
##Lab Exercise 5
Extend courses with a column pgcert which is true if the total is at least 60 credits
```{r}
courses$pgcert <- courses$total >= 60
head(courses)
```

##Lab Exercise 6
Create a data frame with details of all the courses where at least one module is 20 credits or more
```{r}
twentyplus <- subset(courses, courses$credits1>=20 | courses$credits2>=20)
head(twentyplus)

```
##Lab Exercise 7
Create a data frame with details of all the courses which do NOT have databases.
```{r}
noDBs <- subset(courses, courses$module1 != "databases" & courses$module2 != "databases")
head(noDBs)
```
##Lab Exercise 8
Load and read ContactLens.csv
```{r}
lens <- read.csv("ContactLens.csv")
lens
```
##Lab Exercise 9
Show the first 4 rows of the lenses data frame.
```{r}
head(lens, 4)
```
##Lab Exercise 10
Show the last 3 rows of the lenses data frame.
```{r}
tail(lens, 3)
```
##Lab Exercise 11
Check the types of data lenses contains. Does it have any missing values?
```{r}
summary(lens)
```
##Lab Exercise 12
Assign the 3rd instance of lenses age column value pre-presbyopic.
```{r}
lens[3,1] <- "pre-presbyopic"
head(lens)
```
##Lab Exercise 13
Remove column tearproductionRate
```{r}
lens$TearProdRate <- NULL
head(lens)
```
##Lab Exercise 14
Rename column contactLenses to recommendation.
```{r}
colnames(lens)[colnames(lens)=="ContactLens"] <- "recommendation"
head(lens)
```
##Lab Exercise 15
Produce a version of lenses where astigmatism is  numeric (1 if it is a yes and 0 if it has value no) 
```{r}
lens$Astigmatism[lens$Astigmatism == "yes"] <- 1 
lens$Astigmatism[lens$Astigmatism == "no"] <- 0
lens
```



