---
title: "Loan Approval"
author: "Shaibu Abdullateef Topa"
regno: "CST/20/COM/00591"
date: "2025-02-12"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load and inspect the dataset
```{r}
# Load data manipulation library
library(dplyr)

# Read the loan dataset (same folder as this R script)
loan_data <- read.csv("loan_data.csv")

# Show first few rows
head(loan_data)

```
#Summary of dataset
```{r}
# Check dataset structure
str(loan_data)

# Summary statistics of the loan dataset
summary(loan_data)
```
#Handle Missing Values (if any)
```{r}
# Count missing values in each column
colSums(is.na(loan_data))
```

#Convert Categorical Variables - Since KNN works best with numerical data, convert categorical variables into factors.
```{r}
loan_data$person_gender <- as.factor(loan_data$person_gender)

loan_data$person_education <- as.factor(loan_data$person_education)

loan_data$person_home_ownership <- as.factor(loan_data$person_home_ownership)

loan_data$loan_intent <- as.factor(loan_data$loan_intent)

loan_data$previous_loan_defaults_on_file <- as.factor(loan_data$previous_loan_defaults_on_file)

```


#Normalize Numerical Features
```{r}
#Formula
normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}

# Normalize relevant numerical columns
loan_data[, c("person_age", "person_income", "person_emp_exp", "loan_amnt", 
              "loan_int_rate", "loan_percent_income", "cb_person_cred_hist_length", "credit_score")] <- 
  apply(loan_data[, c("person_age", "person_income", "person_emp_exp", "loan_amnt", 
                      "loan_int_rate", "loan_percent_income", "cb_person_cred_hist_length", "credit_score")], 
        2, normalize)

```

#Verify Data to ensure that it is Ready
```{r}
# Check the data again
str(loan_data)

# Check if categorical variables are converted
table(loan_data$loan_status)  # Class distribution

```
#Exploratory Data Analysis - This helps me understand the distribution of values and detect any imbalances or potential issues.
```{r}
# Check structure of loan_data
str(loan_data)

# Summary statistics for numerical variables
summary(loan_data)

# Count unique values for categorical variables
table(loan_data$loan_status)  # Class balance check
table(loan_data$person_gender)
table(loan_data$loan_intent)

```
#Visualizing Data Distributions - Histogram of Numeric Features
```{r}
library(ggplot2)

# Histogram for Age
ggplot(loan_data, aes(x=person_age)) + 
  geom_histogram(fill="blue", bins=30, alpha=0.7) + 
  theme_minimal() +
  ggtitle("Distribution of Age")

# Histogram for Income
ggplot(loan_data, aes(x=person_income)) + 
  geom_histogram(fill="green", bins=30, alpha=0.7) + 
  theme_minimal() +
  ggtitle("Distribution of Income")

```

#Checking Class Imbalance
```{r}
# Class distribution
ggplot(loan_data, aes(x=factor(loan_status), fill=factor(loan_status))) + 
  geom_bar(alpha=0.7) + 
  theme_minimal() +
  ggtitle("Loan Approval Distribution") +
  scale_fill_manual(values = c("red", "green"), labels = c("Rejected", "Approved")) +
  labs(x = "Loan Status", fill = "Status")

```

#KNN can be affected by imbalance because it tends to favor the majority class.
#This means the model might predict 0 (Rejected) too often, leading to #poor recall for 1 (Approved).
#Use ROSE (Random Over-Sampling Examples):
```{r}
library(ROSE)

# Apply oversampling with ROSE
loan_data_balanced <- ROSE(loan_status ~ ., data = loan_data, seed = 123)$data

# Check class balance
table(loan_data_balanced$loan_status)

```


#Verify Balance
```{r}
ggplot(loan_data_balanced, aes(x = factor(loan_status), fill = factor(loan_status))) +
  geom_bar(alpha = 0.7) +
  theme_minimal() +
  ggtitle("Loan Approval Distribution After SMOTE") +
  scale_fill_manual(values = c("red", "green"), labels = c("Rejected", "Approved")) +
  labs(x = "Loan Status", fill = "Status")


```

#Feature Relationships
#Scatter Plot of Loan Amount vs. Income
```{r}
ggplot(loan_data, aes(x=person_income, y=loan_amnt, color=factor(loan_status))) + 
  geom_point(alpha=0.6) + 
  theme_minimal() +
  ggtitle("Income vs Loan Amount by Loan Status")

```

#Training the KNN Model
#Split Data into Training & Testing Sets
#Using Hold Out Estimation
```{r}
# Load required library
library(caTools)

set.seed(123)

# Split the data (80% train, 20% test)
split <- sample.split(loan_data_balanced$loan_status, SplitRatio = 0.8)
train_data <- subset(loan_data_balanced, split == TRUE)
test_data <- subset(loan_data_balanced, split == FALSE)

# Check split result
table(train_data$loan_status)
table(test_data$loan_status)

``` 
#Prepare Data for KNN
#Remove categorical variables and separate labels (target variable):
```{r}
# Load KNN library
library(class)

# Remove categorical columns for KNN (excluding target variable)
train_features <- train_data[, sapply(train_data, is.numeric)]
test_features <- test_data[, sapply(test_data, is.numeric)]

# Extract target labels
train_labels <- train_data$loan_status
test_labels <- test_data$loan_status

```

#Train KNN Model
```{r}
# Train KNN model
set.seed(123)
k_value <- 5
knn_pred <- knn(train = train_features, test = test_features, cl = train_labels, k = k_value)


# Create confusion matrix
library(caret)

# Convert test labels and predictions to factors with same levels
test_labels <- factor(test_labels, levels = unique(train_labels))
knn_pred <- factor(knn_pred, levels = unique(train_labels))

# Create confusion matrix
conf_matrix <- confusionMatrix(knn_pred, test_labels)

# Print results
print(conf_matrix)


```

#Evaluate Model Performance
```{r}
# Check accuracy
accuracy <- mean(knn_pred == test_labels)
print(paste("KNN Model Accuracy:", round(accuracy * 100, 2), "%"))


```
```{r}
print(conf_matrix)

```
