---
title: "Cardio Disease"
author: "Mustapha Shehu Muhammad(CST/19/COM/00332)"
date: "2025-03-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Understanding dataset
age: The person's age in days
sex: The person's sex (1 = women, 2 = male)
height: person's height in cm
weigth: person's weigth in kg
ap_hi: Systolic blood pressure
ap_lo: Diastolic blood pressure
cholestrol: person cholestrol level(1: normal, 2: above normal, 3: well above normal)
glucose: person's glucose level (1: normal, 2: above normal, 3: well above normal)
smoke: whether patient smokes or not(0: NO and 1: YES)
alcohol: Alcohol intake(0: NO and 1: YES)
active: Physical activity(0: NO and 1: YES)
cardio: Heart Diseases---Target Variable(0: NO and 1: YES)

---

## Load Data and Libraries

**Load necessary libraries:**

```{r}

library(RWeka) # For J48 (C4.5 Implementation)
library(C50) # For C5.0
library(caret) # For cross-validation and performance evaluation

```


**Load the dataset:**

```{r}
data <- read.csv("data.csv")
str(data)  # Check the structure

```

**Dataset Summary**

```{r}
summary(data)  # Basic summary

```

**Display First 6 Rows**

```{r}

#displays first 6 rows of the dataset

head(data)

```

## Data Preparation

**Change gender to 0 and 1**

```{r}

# Load dplyr
library(dplyr)

# Map values: 1 → 0, 2 → 1
data$gender <- recode(data$gender, `1` = 0, `2` = 1)

head(data)

```

**Check duplicate rows**
```{r}
# Check for duplicates based on all columns
duplicate_row <- data[duplicated(data), ]

# Print the results
print(duplicate_row)

```

There's no duplicate

**Handle missing values:**

```{r}
# Check total missing values in the dataset
sum(is.na(data))

# Check missing values per column
colSums(is.na(data))


```
There are no missing values

**Drop age_days and rename age_year to age (integer)**

```{r}
# Drop the 'age_days' column
data <- data[, !(names(data) == "age_days")]

# Rename 'age_year' to 'age'
names(data)[names(data) == "age_year"] <- "age"

# Ensure 'age' is an integer
data$age <- as.integer(data$age)

# Verify the changes
head(data)


```

**Drop Id column**

```{r}
# Drop 'id' column
data <- data[, !(names(data) == "id")]

# Verify the changes
head(data)


```

**Outliers Detection**

```{r}
library(ggplot2)
library(gridExtra)  # For side-by-side plots

# Create the two boxplots
plot1 <- ggplot(data, aes(y = ap_hi)) + 
  geom_boxplot() + 
  ggtitle("Boxplot of ap_hi (1)") +
  theme_minimal()

plot2 <- ggplot(data, aes(y = ap_hi)) + 
  geom_boxplot() + 
  ggtitle("Boxplot of ap_hi (2)") +
  theme_minimal()

# Arrange side by side
grid.arrange(plot1, plot2, ncol = 2)

```

clearly we can see that ouliers are present in dataset as such high value of blood-pressure is not possible

** After looking for systolic and diastolic pressure we found that: **
```{r}
# Blood Pressure Categories Data
bp_data <- data.frame(
  Category = c("Normal", "Elevated", "High Blood Pressure (Hypertension) Stage 1", 
               "High Blood Pressure (Hypertension) Stage 2", "Hypertensive Crisis (Consult Doctor Immediately)"),
  Systolic = c("Less than 120", "120 - 129", "130 - 139", "140 or higher", "Higher than 180"),
  Diastolic = c("Less than 80", "Less than 80", "80 - 89", "90 or higher", "Higher than 120")
)

print(bp_data)

```


**Treating Outliers**

```{r}
# Drop rows where systolic pressure (ap_hi) > 230 or diastolic pressure (ap_lo) > 150
data <- data[!(data$ap_hi > 230 | data$ap_lo > 150), ]

# Check the shape (rows and columns)
dim(data)

```
Also we found that a blood pressure reading lower than 90 millimeters of mercury
(mm Hg) for the top number (systolic) or 60 mm Hg for the bottom number
(diastolic) is generally considered low blood pressure.

```{r}

# Drop rows where systolic pressure (ap_hi) < 70 or diastolic pressure (ap_lo) < 55
data <- data[!(data$ap_hi < 70 | data$ap_lo < 55), ]

# Check the shape (rows and columns)
dim(data)

```
```{r}
# Load necessary libraries
library(ggplot2)

# Create the boxplot
ggplot(data, aes(x = factor(cardio), y = height, fill = factor(cardio))) +
  geom_boxplot() +
  scale_fill_manual(values = c("0" = "blue", "1" = "red")) +
  theme_minimal() +
  labs(x = "Cardio", y = "Height", fill = "Cardio") +
  ggtitle("Height by Cardio")

# Calculate IQR, upper limit, and lower limit
Q1 <- quantile(data$height, 0.25, na.rm = TRUE)
Q3 <- quantile(data$height, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1
ul <- Q3 + 1.5 * IQR
ll <- Q1 - 1.5 * IQR

# Print the results
cat("IQR:", IQR, "\n")
cat("Upper limit:", ul, "\n")
cat("Lower limit:", ll, "\n")


```


 outliers present
 
 
```{r}
# Drop rows where height is less than 120 cm
data <- data[data$height >= 120, ]

# Reset row indices (optional, if needed)
rownames(data) <- NULL

# Check the updated data
dim(data)


```

```{r}
# Load necessary libraries
library(ggplot2)

# Create the boxplot with blue and red colors
ggplot(data, aes(x = factor(cardio), y = weight, fill = factor(cardio))) +
  geom_boxplot() +
  scale_fill_manual(values = c("0" = "blue", "1" = "red")) +
  theme_minimal() +
  labs(x = "Cardio", y = "Weight", fill = "Cardio") +
  ggtitle("Weight by Cardio")

# Calculate IQR, upper limit, and lower limit
Q1 <- quantile(data$weight, 0.25, na.rm = TRUE)
Q3 <- quantile(data$weight, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1
ul <- Q3 + 1.5 * IQR
ll <- Q1 - 1.5 * IQR

# Print the results
cat("IQR:", IQR, "\n")
cat("Upper limit:", ul, "\n")
cat("Lower limit:", ll, "\n")


```

Outlier detected

```{r}
# Drop rows where weight is greater than 180
data <- data[data$weight <= 180, ]

# Reset row indices (optional, if needed)
rownames(data) <- NULL

# Check the updated data
dim(data)


```

```{r}
# Drop rows where weight is less than 30
data <- data[data$weight >= 30, ]

# Reset row indices (optional, if needed)
rownames(data) <- NULL

# Check the updated data
head(data)

```




## Exploratory Data Analysis (EDA)

**Visualize feature distributions:**

```{r, fig.height=16, fig.width=16}

# Load ggplot2 library
library(ggplot2)

# Set plot layout to 2 plots per row
gridExtra::grid.arrange(
  grobs = lapply(names(data), function(col) {
    ggplot(data, aes(x = .data[[col]], fill = as.factor(cardio))) +
      geom_histogram(color = "red", bins = 30, alpha = 0.7) +
      labs(title = col, fill = "cardio") +
      theme_minimal()
  }),
  ncol = 2
)


```
Key Insights:
Blood pressure, cholesterol, and glucose levels seem to be the most telling
indicators of cardiovascular disease in this dataset. Age and weight also show
noticeable trends. Lifestyle factors don’t show a strong correlation visually,
but could still play a role when combined with other features.

**UNIVARIATE ANALYSIS**

```{r}
library(ggplot2)

ggplot(data, aes(x = age, fill = as.factor(cardio))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("0" = "green", "1" = "red"), labels = c("NO", "YES")) +
  labs(x = "Age", fill = "Cardio") +
  theme_minimal()


```

```{r}
library(dplyr)
library(psych)

# Group by 'cardio' and summarize 'age'
data %>%
  group_by(cardio) %>%
  summarise(
    count = n(),
    mean = mean(age, na.rm = TRUE),
    sd = sd(age, na.rm = TRUE),
    min = min(age, na.rm = TRUE),
    Q1 = quantile(age, 0.25, na.rm = TRUE),
    median = median(age, na.rm = TRUE),
    Q3 = quantile(age, 0.75, na.rm = TRUE),
    max = max(age, na.rm = TRUE)
  )


```
**Observation:**

Person's suffering from heart related issue tend to have higher age than other
ones. As we can see here that avg age of person having heart disease is
54.43 and others have **51.21**
As per the plot we can see that **pdf curve of person having heart disease is
left skewed means person with lower age have less chance of having heart
problems.



```{r}

library(ggplot2)

# Create the plot
ggplot(data, aes(x = weight, fill = factor(cardio))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("0" = "green", "1" = "red"), labels = c("NO", "YES")) +
  labs(x = "Weight", y = "Density", fill = "Heart Disease") +
  theme_minimal() +
  theme(legend.position = "top")

```
```{r}
library(dplyr)

# Group by cardio and summarize weight
data %>%
  group_by(cardio) %>%
  summarize(
    count = n(),
    mean = mean(weight, na.rm = TRUE),
    std = sd(weight, na.rm = TRUE),
    min = min(weight, na.rm = TRUE),
    q1 = quantile(weight, 0.25, na.rm = TRUE),
    median = median(weight, na.rm = TRUE),
    q3 = quantile(weight, 0.75, na.rm = TRUE),
    max = max(weight, na.rm = TRUE)
  )



```
**Observation:**

person having more weight have higher chance of having heart diseases. As we can
see from the graphs and also from the statistical data that Avg. weight of
person having heart problems(76.91) is more than those who don't have.


```{r}
library(ggplot2)

ggplot(data, aes(x = height, fill = as.factor(cardio))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("0" = "green", "1" = "red"), labels = c("NO", "YES")) +
  labs(x = "Height", fill = "Heart Disease") +
  theme_minimal()


```

```{r}
library(dplyr)

# Group by 'cardio' and summarize 'height'
data %>%
  group_by(cardio) %>%
  summarise(
    count = n(),
    mean = mean(height, na.rm = TRUE),
    std = sd(height, na.rm = TRUE),
    min = min(height, na.rm = TRUE),
    q1 = quantile(height, 0.25, na.rm = TRUE),
    median = median(height, na.rm = TRUE),
    q3 = quantile(height, 0.75, na.rm = TRUE),
    max = max(height, na.rm = TRUE)
  )


```

**Observation:**

Height is doesn't play any role in determining heart diseases. As we can see
both the plot overlaps and avg height of person with heart problem and without
heart problems is also same.

**Blood pressure analysis**


```{r}
library(ggplot2)

ggplot(data, aes(x = ap_hi, fill = as.factor(cardio))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("0" = "green", "1" = "red"), labels = c("NO", "YES")) +
  labs(x = "Blood Pressure", fill = "Heart Disease") +
  theme_minimal()


```

```{r}
library(dplyr)

data %>%
  group_by(cardio) %>%
  summarise(
    count = n(),
    mean = mean(ap_hi, na.rm = TRUE),
    std = sd(ap_hi, na.rm = TRUE),
    min = min(ap_hi, na.rm = TRUE),
    q1 = quantile(ap_hi, 0.25, na.rm = TRUE),
    median = median(ap_hi, na.rm = TRUE),
    q3 = quantile(ap_hi, 0.75, na.rm = TRUE),
    max = max(ap_hi, na.rm = TRUE)
  )


```
```{r}
library(ggplot2)

ggplot(data, aes(x = ap_lo, fill = factor(cardio))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("0" = "green", "1" = "red"), labels = c("NO", "YES")) +
  labs(x = "Blood Pressure", fill = "Heart Disease") +
  theme_minimal() +
  theme(legend.title = element_blank())


```
```{r}

data %>%
  group_by(cardio) %>%
  summarise(
    count = n(),
    mean = mean(ap_lo, na.rm = TRUE),
    std = sd(ap_lo, na.rm = TRUE),
    min = min(ap_lo, na.rm = TRUE),
    q1 = quantile(ap_lo, 0.25, na.rm = TRUE),
    median = median(ap_lo, na.rm = TRUE),
    q3 = quantile(ap_lo, 0.75, na.rm = TRUE),
    max = max(ap_lo, na.rm = TRUE)
  )

```
**Observation:**

As we can see that people who have high blood pressure have high chance that
they have heart disease. Avg. Blood pressure of person having heart problems is
134.07 is more than the normal level.


**BIVARIATE ANALYSIS**

```{r}
library(ggplot2)

ggplot(data, aes(x = ap_lo, y = ap_hi, color = factor(cardio), shape = factor(cardio))) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("0" = "lightblue", "1" = "red")) +
  labs(x = "Diastolic Pressure (ap_lo)", y = "Systolic Pressure (ap_hi)", color = "Cardio", shape = "Cardio") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  theme(legend.position = "top") +
  ggtitle("Systolic vs Diastolic Blood Pressure by Cardio Condition")


```


**Observation:**
People with heart problems have higher systolic pressure and diastolic
pressure values as per the plot.


**Smoke and alcohol analysis**

```{r}
library(dplyr)

data %>%
  group_by(smoke) %>%
  summarise(
    ap_hi_count = n(),
    ap_hi_mean = mean(ap_hi, na.rm = TRUE),
    ap_hi_sd = sd(ap_hi, na.rm = TRUE),
    ap_hi_min = min(ap_hi, na.rm = TRUE),
    ap_hi_q1 = quantile(ap_hi, 0.25, na.rm = TRUE),
    ap_hi_median = median(ap_hi, na.rm = TRUE),
    ap_hi_q3 = quantile(ap_hi, 0.75, na.rm = TRUE),
    ap_hi_max = max(ap_hi, na.rm = TRUE),
    ap_lo_mean = mean(ap_lo, na.rm = TRUE),
    ap_lo_sd = sd(ap_lo, na.rm = TRUE),
    ap_lo_min = min(ap_lo, na.rm = TRUE),
    ap_lo_q1 = quantile(ap_lo, 0.25, na.rm = TRUE),
    ap_lo_median = median(ap_lo, na.rm = TRUE),
    ap_lo_q3 = quantile(ap_lo, 0.75, na.rm = TRUE),
    ap_lo_max = max(ap_lo, na.rm = TRUE)
  )


```
```{r}
library(dplyr)

data %>%
  group_by(alco) %>%
  summarise(
    ap_hi_count = n(),
    ap_hi_mean = mean(ap_hi, na.rm = TRUE),
    ap_hi_sd = sd(ap_hi, na.rm = TRUE),
    ap_hi_min = min(ap_hi, na.rm = TRUE),
    ap_hi_q1 = quantile(ap_hi, 0.25, na.rm = TRUE),
    ap_hi_median = median(ap_hi, na.rm = TRUE),
    ap_hi_q3 = quantile(ap_hi, 0.75, na.rm = TRUE),
    ap_hi_max = max(ap_hi, na.rm = TRUE),
    ap_lo_mean = mean(ap_lo, na.rm = TRUE),
    ap_lo_sd = sd(ap_lo, na.rm = TRUE),
    ap_lo_min = min(ap_lo, na.rm = TRUE),
    ap_lo_q1 = quantile(ap_lo, 0.25, na.rm = TRUE),
    ap_lo_median = median(ap_lo, na.rm = TRUE),
    ap_lo_q3 = quantile(ap_lo, 0.75, na.rm = TRUE),
    ap_lo_max = max(ap_lo, na.rm = TRUE)
  )


```
**Observation:**
Although there is not much difference but still we can say that person who
smoke or take alcohol have higher blood pressure than other person. And hence
chances that they can have heart problems also increases.

Cardio Vs Smoke and Cardio Vs Alco
```{r}

library(ggplot2)
library(gridExtra)

# Plot 1: Cardio vs Smoke
plot1 <- ggplot(data, aes(x = as.factor(cardio), fill = as.factor(smoke))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("blue", "red")) +
  labs(x = "Cardio", fill = "Smoke") +
  theme_minimal()

# Plot 2: Cardio vs Alco
plot2 <- ggplot(data, aes(x = as.factor(cardio), fill = as.factor(alco))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("blue", "red")) +
  labs(x = "Cardio", fill = "Alco") +
  theme_minimal()

# Arrange side by side
grid.arrange(plot1, plot2, ncol = 2)

```

**Observation:**
plot shows that those who smoke and have high blood pressure are mostly having
heart problems


**Cholesterol analysis**


```{r}
library(dplyr)

# Group by cholesterol and summarize ap_hi and ap_lo
data %>%
  group_by(cholesterol) %>%
  summarise(
    ap_hi_mean = mean(ap_hi, na.rm = TRUE),
    ap_hi_sd = sd(ap_hi, na.rm = TRUE),
    ap_hi_min = min(ap_hi, na.rm = TRUE),
    ap_hi_max = max(ap_hi, na.rm = TRUE),
    ap_lo_mean = mean(ap_lo, na.rm = TRUE),
    ap_lo_sd = sd(ap_lo, na.rm = TRUE),
    ap_lo_min = min(ap_lo, na.rm = TRUE),
    ap_lo_max = max(ap_lo, na.rm = TRUE)
  )


```

```{r}
library(ggplot2)

# Scatter plot with ap_lo on x-axis, ap_hi on y-axis, cardio as shape, and cholesterol as color
ggplot(data, aes(x = ap_lo, y = ap_hi, shape = as.factor(cardio), color = as.factor(cholesterol))) +
  geom_point() +
  scale_color_manual(values = c("orange", "red", "purple")) +
  labs(x = "ap_lo", y = "ap_hi", shape = "Cardio", color = "Cholesterol") +
  theme_minimal() +
  theme(legend.position = "right") +
  ggtitle("Scatterplot of Blood Pressure with Cardio and Cholesterol")


```
**Observation:**
person with high cholesterol level have higher blood pressure and have more
chance to have heart problems

```{r}
library(ggplot2)

# Count plot for cholesterol with cardio as hue
ggplot(data, aes(x = as.factor(cholesterol), fill = as.factor(cardio))) +
  geom_bar(position = "dodge") +
  labs(x = "Cholesterol", fill = "Cardio") +
  theme_minimal() +
  ggtitle("Cholesterol vs Cardio Count Plot") +
  theme(plot.title = element_text(hjust = 0.5))


```


**Observation:**
Most of the Person with **cholesterol level 2 and 3** have  heart disease.

**Gender analysis**

```{r}

library(ggplot2)

# Count plot for cardio with gender as hue
ggplot(data, aes(x = as.factor(cardio), fill = as.factor(gender))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("lightgreen", "red")) +
  labs(x = "Cardio", fill = "Gender") +
  theme_minimal() +
  ggtitle("Cardio vs Gender Count Plot") +
  theme(plot.title = element_text(hjust = 0.5))

```


**Observation:**

Over 20000 females and 10000 males have heart diseases.

**Glucose and Physical Activity**

```{r}
library(ggplot2)
library(gridExtra)

# Gluc vs Cardio Count Plot
p1 <- ggplot(data, aes(x = as.factor(gluc), fill = as.factor(cardio))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("lightgreen", "red")) +
  labs(x = "Gluc", fill = "Cardio") +
  theme_minimal() +
  ggtitle("Gluc vs Cardio") +
  theme(plot.title = element_text(hjust = 0.5))

# Active vs Cardio Count Plot
p2 <- ggplot(data, aes(x = as.factor(active), fill = as.factor(cardio))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("lightgreen", "red")) +
  labs(x = "Active", fill = "Cardio") +
  theme_minimal() +
  ggtitle("Active vs Cardio") +
  theme(plot.title = element_text(hjust = 0.5))

# Arrange plots side by side
grid.arrange(p1, p2, ncol = 2)


```

**Observation**:
Both the feature doesn't seem to have much correlation with cardiac problems.
Hence these features are not that important as others.


**Correlation Matrix**

```{r}
library(corrplot)

# Calculate the correlation matrix
corr_matrix <- cor(data)

# Plot the heatmap
corrplot(corr_matrix, method = "color", addCoef.col = "black", tl.col = "black", 
         tl.srt = 45, col = colorRampPalette(c("red", "white", "blue"))(200), 
         number.cex = 0.7)

```

## Modelling 

**Convert target variable to factor (if not already)**

```{r}
data$cardio <- as.factor(data$cardio)

# Split dataset into training (80%) and testing (20%)
set.seed(123)
trainIndex <- createDataPartition(data$cardio, p = 0.8, list = FALSE)
trainData <- data[trainIndex, ]
testData <- data[-trainIndex, ]

dim(trainData)
dim(testData)


```


**Apply J48 (C4.5 Implementation in RWeka)**

```{r}
# Train J48 model
library(RWeka)
library(caret)

j48_model <- J48(cardio ~ ., data = trainData)

# Predict on test set
j48_pred <- predict(j48_model, testData)

# Evaluate Model Performance
conf_matrix <- confusionMatrix(j48_pred, testData$cardio)

# Print confusion matrix
print(conf_matrix)

# Extract metrics
precision <- conf_matrix$byClass["Pos Pred Value"]  # Precision
recall <- conf_matrix$byClass["Sensitivity"]        # Recall
f1_score <- 2 * (precision * recall) / (precision + recall)  # F1-Score

# Display metrics
cat("Precision:", precision, "\n")
cat("Recall:", recall, "\n")
cat("F1-Score:", f1_score, "\n")


```
**Apply C5.0 Decision Tree**

```{r}
# Load necessary libraries
library(C50)
library(caret)

# Train C5.0 model
c50_model <- C5.0(cardio ~ ., data = trainData)

# Predict on test set
c50_pred <- predict(c50_model, testData)

# Evaluate Model Performance
conf_matrix <- confusionMatrix(c50_pred, testData$cardio)

# Print confusion matrix
print(conf_matrix)

# Extract metrics
precision <- conf_matrix$byClass["Pos Pred Value"]  # Precision
recall <- conf_matrix$byClass["Sensitivity"]        # Recall
f1_score <- 2 * (precision * recall) / (precision + recall)  # F1-Score

# Display metrics
cat("Precision:", precision, "\n")
cat("Recall:", recall, "\n")
cat("F1-Score:", f1_score, "\n")


```
**Apply C5.0 Rules-Based Model**

```{r}

# Load necessary libraries
library(C50)
library(caret)

# Train C5.0 model with rule-based classifier
c50_rules_model <- C5.0(cardio ~ ., data = trainData, rules = TRUE)

# Predict on test set
c50_rules_pred <- predict(c50_rules_model, testData)

# Evaluate Model Performance
conf_matrix <- confusionMatrix(c50_rules_pred, testData$cardio)

# Print confusion matrix
print(conf_matrix)

# Extract metrics
precision <- conf_matrix$byClass["Pos Pred Value"]  # Precision
recall <- conf_matrix$byClass["Sensitivity"]        # Recall
f1_score <- 2 * (precision * recall) / (precision + recall)  # F1-Score

# Display metrics
cat("Precision:", precision, "\n")
cat("Recall:", recall, "\n")
cat("F1-Score:", f1_score, "\n")

```
**Apply Cross-Validation**

```{r}
# Set seed for reproducibility
set.seed(123)

# Define cross-validation method (10-fold CV)
train_control <- trainControl(method = "cv", number = 10)

# Train the model using cross-validation
model <- train(
  cardio ~ .,               # cardio is the target variable
  data = data,              # your dataset
  method = "rpart",         # Decision Tree (you can use other methods)
  trControl = train_control
)

# Print model summary
print(model)

# Check model performance
model$results

```

**Accuracy comparison**

```{r}

j48_acc <- mean(j48_pred == testData$cardio)
c50_acc <- mean(c50_pred == testData$cardio)
c50_rules_acc <- mean(c50_rules_pred == testData$cardio)

accuracy_results <- data.frame(
  Model = c("J48", "C5.0 Tree", "C5.0 Rules"),
  Accuracy = c(j48_acc, c50_acc, c50_rules_acc)
)

print(accuracy_results)

```

# Model Comparison

Here i reshape the data for visualization for the model comparison
```{r}
# Load libraries
library(ggplot2)
library(reshape2)

# Create a data frame for your model metrics
model_comparison <- data.frame(
  Model = c("J48", "C5.0 Tree", "C5.0 Rules"),
  Accuracy = c(0.728, 0.733, 0.733),
  Precision = c(0.716, 0.716, 0.716),
  Recall = c(0.781, 0.781, 0.781),
  F1_Score = c(0.747, 0.747, 0.747)
)

# Reshape data for visualization
model_comparison_long <- melt(model_comparison, id.vars = "Model")

# Plot performance metrics
ggplot(model_comparison_long, aes(x = Model, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Cardio Model Performance Comparison", x = "Decision Tree Model", y = "Metric Value") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red", "green", "purple"))

```

Comments:
The C5.0 Rules model performed marginally better than the other two. The
difference is very small, suggesting that the models are learning similar
patterns in the data. Accuracy around 72–73% is decent,

These metrics indicate a reasonably balanced model performance:

Precision (0.716): About 71.6% of the positive predictions were accurate.
Recall (0.781): The model correctly identified 78.1% of actual positives.
F1-Score (0.747): This value balances precision and recall, showing an overall performance of 74.7%.

